Using Local PCA to Investigate Population Structure Along the Genome

### 2016-05-05 medicago/local_pca_on_medicago.Rmd
rewrote eigen_windows to take functions

@@

The file of sample info was obtained from the HTML table at [http://www.medicagohapmap.org/hapmap/germplasm](http://www.medicagohapmap.org/hapmap/germplasm).
To check this contains information about all samples,
first get the list of samples in the .bcf files:
```
bcftools query -l chr1-filtered-set-2014Apr15.bcf > samples_in_bcf.txt
```
then compare to the table of information:
```{r sample_info}
samples <- read.table("sample_info.tsv", sep="\t", header=TRUE)
bcf.samples <- scan("samples_in_bcf.txt", what='char')
setdiff( bcf.samples , samples$ID )
```
For some reason, the bcf files refer to `HM020-I` while in the the sample info file there is `HM020`.
We'll assume they are the same (but keep an eye on that one!).
Here's where they're all from:
```{r show_sample_info}
bcf.samples <- gsub("-I","",bcf.samples)
samples <- droplevels(subset(samples, ID %in% bcf.samples))
sort( table(samples$Country.of.Origin) )
```

@@

To get windows out of the VCF file and recode them, we'll just pipe a call to `bcftools`.
Here's a dummy version that will pull out windows of 100 SNPs for only the first four individuals.
```{r win_fn}
get.indivs <- "'HM001,HM002,HM003,HM004'"
bcf.file <- "chr1-filtered-set-2014Apr15.bcf"
# first read in the info about sites
bcf.con <- pipe(paste("bcftools query -f '%CHROM\\t%POS\\n' -r chr1:1-100000",bcf.file),open="r")
bcf.sites <- read.table(bcf.con, sep='\t')
colnames(bcf.sites) <- c("chrom","pos")
close(bcf.con)
# now use bcftools to pull out sites we want
#' This gives a string of the form "chr:start-end,chr:start-end" as required by bcftools
get_regions <- function (sites) {
    if (nrow(sites)==0) { return( NULL ) }
    pos <- as.numeric( sites[,2] )
    do.call( paste, c( tapply( 1:nrow(sites), sites[,1], function (kk) {
                            paste0( sites[kk[1],1], ":", min(pos[kk]), "-", max(pos[kk]) )
                        } ), list(sep=",") ) )
}
#' This will pull out bits of the bcf file corresponding to a regions string and convert it to numeric
get_sites <- function (regions) {
    gt.text <- data.table::fread( paste( c("bcftools", "query", "-f", "'[ %GT]\\n'", "-r", regions, "-s", get.indivs, bcf.file), collapse=" " ), header=FALSE, sep=' ', data.table=FALSE )
    gt <- c(0,1,1,2)[match( unlist(gt.text), c("0/0","0/1","1/0","1/1") )]
    dim(gt) <- dim(gt.text)
    return(gt)
}
win.fn <- function (n,win.snps=100) {
    sites <- bcf.sites[seq(1+(n-1)*win.snps,length.out=win.snps),]
    regions <- get_regions(sites)
    get_sites(regions)
}
win.fn(20,10)  # get the 20th window of 10 SNPs
```

Here's an alternate version that pulls out windows based on basepairs.
Note that this version assumes we have only one chromosome, starting at zero.
```{r win_fn_bp}
win.fn.bp <- function (n,win.bp=200) {
    win.start <- 1+(n-1)*win.bp
    win.end <- win.start + win.bp-1
    regions <- get_regions( cbind( chrom="chr1", pos=c(win.start,win.end) ) )
    get_sites(regions)
}
win.fn.bp(20,50)
```

@@

>>>>

```{r setup}
devtools::load_all("../package")
```

@@

The file of sample info was obtained from the HTML table at [http://www.medicagohapmap.org/hapmap/germplasm](http://www.medicagohapmap.org/hapmap/germplasm).
To check this contains information about all samples,
first get the list of samples in the .bcf files:
```
bcftools query -l chr1-filtered-set-2014Apr15.bcf > samples_in_bcf.txt
```
then compare to the table of information:
```{r sample_info}
samples <- read.table("data/sample_info.tsv", sep="\t", header=TRUE)
bcf.samples <- scan("data/samples_in_bcf.txt", what='char')
setdiff( bcf.samples , samples$ID )
```
For some reason, the bcf files refer to `HM020-I` while in the the sample info file there is `HM020`.
We'll assume they are the same (but keep an eye on that one!).
Here's where they're all from:
```{r show_sample_info}
bcf.samples <- gsub("-I","",bcf.samples)
samples <- droplevels(subset(samples, ID %in% bcf.samples))
sort( table(samples$Country.of.Origin) )
```

@@

To get windows out of the VCF file and recode them, we'll just pipe a call to `bcftools`.
Here's a dummy version that will pull out windows of 100 SNPs for only the first four individuals.
```{r win_fn}
get.indivs <- c("HM001","HM002","HM003","HM004")
bcf.file <- "data/chr1-filtered-set-2014Apr15.bcf"
# first read in the info about sites
bcf.con <- pipe(paste("bcftools query -f '%POS\\n' -r chr1:1-100000",bcf.file),open="r")
bcf.sites <- list( chr1=scan(bcf.con) )
close(bcf.con)
# now use bcftools to pull out sites we want
win.fn <- function (n,win.snps=100) {
    sites <- list( chr1=bcf.sites[[1]][seq(1+(n-1)*win.snps,length.out=win.snps)] )
    regions <- region_string(sites)
    query_genotypes( bcf.file, regions, get.indivs )
}
win.fn(20,10)  # get the 20th window of 10 SNPs
```

Here's an alternate version that pulls out windows based on basepairs.
Note that this version assumes we have only one chromosome, starting at zero.
```{r win_fn_bp}
win.fn.bp <- function (n,win.bp=200) {
    win.start <- 1+(n-1)*win.bp
    win.end <- win.start + win.bp-1
    regions <- paste0( "chr1", ":", win.start, ",", win.end )
    query_genotypes( bcf.file, regions, get.indivs )
}
win.fn.bp(20,50)
```

@@

## Doing this for everything

Now let's see how it goes to do this for an entire chromosome.
On chromosome 1 there are about 5 million SNPs, so we'll do it in windows of 5,000.
First, windows by SNPs.
```{r lpca_chr1}
sites <- vcf_positions(bcf.file)
win.fn.snp <- vcf_windower(bcf.file, size=5e3, type="snp", sites=sites, samples=get.indivs) 
```
